Using Fundamental Measure Theory to Treat the Correlation Function of the Inhomogeneous Hard-Sphere Fluid

### 2012-08-29 papers/contact/paper.tex
edits on paper Tests skipped!

\begin{figure}
\includegraphics[width=5cm]{figs/contact}
\caption{An illustration of the prospect of changing the radius of
  hard spheres only where they contact a given point, in order to find
  the pressure on those spheres at just that point. }
\label{fig:contact}
\end{figure}

We now address the symmetrically averaged correlation function, which
is defined in Equation~\ref{eq:gS}.  This corresponds to the
correlation function averaged for spheres \emph{touching at a given
  point}.  It is a symmetric measure, since the two spheres involved
are both treated in the same way.  The density of spheres that could
be touching at the point $\rr$ is precisely the density $n_0(\rr)$
from FMT (see Figure~\ref{fig:n0} for an illustration $n_0$).  Since
$n_0$ is the fundamental measure of sphere surface contact, it makes
sense to define a corresponding correlation function $g_\sigma^S$.
This is closely related what is done by Yu and
Wu\cite{yu2002fmt-dft-inhomogeneous-associating}, as will be discussed
in Section~\ref{sec:yuwu}.

The procedure for finding the symmetrically averaged correlation
function $g_\sigma^S(\rr)$ is less straight-forward than that of the
previous section, since we must find the average pressure exerted on
the portion of the surface of spheres which is at position $\rr$.  We
achieve this, as in the previous section, by defining a
position-dependent hard-sphere diameter $\sigma(\rr)$, but this time
we define the diameter to be dependent on the position at which the
fundamental measure itself is evaluated:
\begin{align}
  n_3^S(\rr) &= \int n(\rr')\Theta(|\rr-\rr'|-\sigma(\rr)/2)d\rr'
  \label{eq:n3S}
\end{align}
This has the effect of expanding spheres as they touch the point
$\rr$, a concept illustrated in Figure~\ref{fig:contact}.This
enables a similar derivation to that which gave us $g_\sigma^A(\rr)$:
\begin{align}
  p_{HS}(\mathbf{r}) &= \frac{1}{n_0(\rr) 2\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta \sigma(\mathbf{r})} \label{eq:p_{HS}^S}\\
%  \ncontact(\rr) &= \frac{1}{n_0(\rr) k_BT 2\pi \sigma^2} \frac{\delta
%    A_{HS}}{\delta \sigma(\mathbf{r})} \\
  g_\sigma^S(\rr)% &= \frac{\ncontact(\rr)}{n_0(\rr)} \\
  &= \frac{1}{n_0(\rr)^2}\frac{1}{ k_BT 2\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta \sigma(\mathbf{r})} \label{eq:g-A-exact}
\end{align}
where the functional derivative $\frac{\delta A_{HS}}{\delta
  \sigma(\mathbf{r})}$ in the above formula is computed with the
symmetric formulation of the fundamental measures, as in
Equation~\ref{eq:n3S}.  Details of some aspects of finding this
derivative using FMT are given in Appendix~\ref{appendix:g-S}.  The
equation for $g_\sigma^S$ is computationally simpler than that for
$g_\sigma^A$, as it only involves convolutions of the density itself,
although it does require two new fundamental measures
\begin{align}
  n_{2'}(\rr) &= \int n(\rr') \delta'(\left|\rr - \rr'\right| -
  \sigma/2) d\rr' \label{n2p} \\
  \mathbf{n}_{2V'}(\rr) &=
  \int n(\rr') \delta'(\left|\rr - \rr'\right| - \sigma/2)
  \frac{\rr-\rr'}{|\rr-\rr'|}d\rr'
  \label{n2vp}
\end{align}
where $\delta'$ is the derivative of the Dirac delta function.  These
convolutions are easily expressed in reciprocal space, and can be
easily and efficiently computed.

\subsection{Gross's asymmetrically averaged correlation functional}\label{sec:gross}
One approximation for the correlation function is that of
Gross\cite{gross2009density}, which is of the asymmetrically averaged
variety ($g_\sigma^A$):
\begin{align}
  g_\sigma^\text{Gross,A}(\rr) &= \frac{1 - \frac{\pi}{12}\sigma^3n_A(\rr)}{\left(1 -
    \frac{\pi}{6}\sigma^3n_A(\rr)\right)^3}
\end{align}
where $n_A$ is the averaged density defined in Equation~\ref{eq:nA}.
This formula is arrived at by using the density averaged over all
spheres that could be touching a sphere at point $\rr$ in the
Carnahan-Starling equation for the correlation function at contact,
given in Equation~\ref{eq:cs-g}.

@@

@@

We performed a Monte-Carlo simulation of the hard sphere fluid to
measure the contact density as a function of position for a few simple
geometries.  For each scenario, we compute and compare several
quantities.  We compare the Monte Carlo density with the DFT
prediction.  We also plot Monte Carlo results for the two contact
densities: the contact density evaluated at contact, and the contact
density evaluated at the sphere location.  Alongside these, we plot
the contact density of Yu and
Wu~\cite{yu2002fmt-dft-inhomogeneous-associating}, the contact density
of Gross~\cite{gross2009density}, and the ``contact density at
sphere'' from Section~\ref{sec:g-A}.  In principle, the
functional of Yu (FIXME) should match the CenConDensity, while the DFT
at sphere should match the ConDensity...

\subsection{Hard spheres at a hard wall}

We begin by examining the simplest situation with an inhomogeneous
density:  the case of hard spheres near a single hard wall.  This is
shown in Figures~\ref{fig:walls-10}-\ref{fig:walls-40}.

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/walls-10}
  \caption{Density and contact densities at a wall with bulk filling
    fraction of 0.1.}
  \label{fig:walls-10}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/inner-4-10}
  \caption{Density and contact densities at a xxx with bulk filling
    fraction of 0.1.}
  \label{fig:inner-4-10}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/inner-12-10}
  \caption{Density and contact densities at a xxx with bulk filling
    fraction of 0.1.}
  \label{fig:inner-12-10}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/outer-16-01}
  \caption{Density and contact densities at a xxx with bulk filling
    fraction of 0.1.}
  \label{fig:outer-10}
\end{figure}

\subsection{Hard spheres confined in a spherical cavity}

The next geometry we consider is a number of hard spheres confined
within a spherical cavity.  In
Figures~\ref{fig:sphere-8}-\ref{fig:sphere-16}, we show the densities
and contact densities for several cavity sizes \fixme{at a variety of
  filling fractions}.  We chose cavities that are large enough that
the center of the sphere is bulk-like, so that we can interpret the
behavior near the interface as behavior that should be typical of the
behavior of the hard sphere fluid near a concave surface.

\subsection{Hard spheres around a hard sphere}

It is also interesting to study the hard sphere fluid at a convex
surface.  We did so by placing a (larger) hard sphere in the
hard-sphere fluid, to see how the correlation function behaves near
the surface of that hard sphere.  This is particularly relevant when
we consider using classical density-functional theory to model
solvation.

\section{Results}

%% \subsection{Test-particle result}

%% Here we discuss comparison of the correlation function calculated via
%% both Monte Carlo and the test-particle approach (see
%% Figures~\ref{fig:inner-4-10} and~\ref{fig:inner-4-40}).

@@

Here we discuss the low-density results (filling fraction of 0.1),
which are shown in Figures~\ref{fig:walls-10}-\ref{fig:outer-10}.

@@

We must therefore compute both the $\frac{\delta A_{HS}}{\delta
  n(\mathbf{r}')}$, etc., terms and the $\frac{\delta
  n(\mathbf{r}')}{\delta \sigma(\mathbf{r})}$, etc., terms.  We will
compute the $\frac{\delta A_{HS}}{\delta n(\mathbf{r}')}$, etc., terms
for both the functional derived in the White Bear I formulation and
for that in the White Bear II formulation.  Following this we will
compute the $\frac{\delta n(\mathbf{r}')}{\delta \sigma(\mathbf{r})}$,
etc., terms for both the $A$ and $S$ cases.  These latter derivitives will
differ for the $A$ and $S$ cases according to the qualitive concepts
described above.

@@

The derivatation of the asymmetrically averaged correlation function
$g_\sigma^A(\rr)$ requires that we express the free energy in terms of
a set of weighted densities in which the radius of the weighting
kernel is determined according to the position of the sphere in
question---which is to say, the location where the density itself is
evaluated.  This is the most natural way to construct such a
functional, in which the radius of spheres is a function of position.
Thus we define $n_3^A$ as
\begin{align}
  n_3^{A}(\rr') &= \int n(\rr'') \Theta(\left|\rr' - \rr''\right| -\sigma(\rr'')/2) d\rr''
\end{align}
and we define the other weighted densities similarly.  In order to
compute derivatives of the free energy with respect to $R(\rr)$, we
need to find $g_\sigma^A$.  In the case of $n_3^A$, we need
\begin{align}
  \frac{\delta n_3^{A} (\rr')}{\delta \sigma(\rr)} &=
  \frac 12 \int n (\rr'') \delta(|\rr' - \rr''| - \sigma(\rr'')/2) \delta(\rr-\rr'') d\rr'' \\
   &= n (\rr) \delta(|\rr' - \rr| - \sigma(\rr)/2)
\end{align}
Similarly, for $n_2^A$, we find
\begin{align}
  n_2^{A}(\rr') &= \int n(\rr'') \delta(|\rr' - \rr''| - \sigma(\rr'')/2) d \rr''\\
  \frac{\delta n_2^{A}(\rr')}{\delta \sigma(\rr)} &= -\frac 12 n(\rr) \delta'(|\rr' - \rr| - \sigma(\rr)/2)\\
\end{align}
This is somewhat more challenging, since it involves an unfamiliar
function, the derivative the Dirac delta function $\delta'$.
%
The derivatives of the remaining scalar densities can be reduced to
sums of the terms above:
\begin{align}
  n_1(\rr') &= \int \mathbf{dr''} \frac{n(\rr'')}{2\pi \sigma(\rr'')}
  \delta(|\rr'-\rr''| - \sigma(\rr'')/2) \\
  \frac{\delta n_1(\rr')}{\delta \sigma(\rr)}
  &= -\frac{n(\rr)}{4\pi
    \sigma(\rr)}\delta'(|\rr'-\rr| - \sigma(\rr)/2)
  -
  \frac{n(\rr)}{2\pi
    \sigma(\rr)^2}\delta(|\rr'-\rr| - \sigma(\rr)/2)
\end{align}
and
\begin{align}
  n_0(\rr') &= \int \mathbf{dr''} \frac{n(\rr'')}{\pi \sigma(\rr'')^2}
  \delta(|\rr'-\rr''| - \sigma(\rr'')/2) \\
  \frac{\delta n_0(\rr')}{\delta \sigma(\rr)}
  &= -\frac{n(\rr)}{2\pi
    \sigma(\rr)^2}\delta'(|\rr'-\rr| - \sigma(\rr)/2)
  -
  2 \frac{n(\rr)}{\pi
    \sigma(\rr)^3}\delta(|\rr'-\rr| - \sigma(\rr)/2)
\end{align}
The vector-weighted densities have no additional occurrences of $R$,
their functional derivatives come out similarly:
\begin{align}
  \mathbf{n}_{V2}^{A}(\rr') &= \int n(\rr') \delta(|\rr' - \rr''| - \sigma(\rr'')/2)
    \frac{\rr'-\rr''}{|\rr'-\rr''|} d \rr''\\
  \frac{\delta \mathbf{n}_{V2}^{A}(\rr')}{\delta \sigma(\rr)} &= \frac 12 n(\rr) \delta'(|\rr' - \rr| - \sigma(\rr)/2)
    \frac{\rr-\rr'}{|\rr-\rr'|}\\
\end{align}
and analogously for $\mathbf{n}_{V1}^A$:
\begin{align}
  \mathbf{n}_{V1}^A(\rr') &= \int d\rr'' \frac{n(\rr'')}{2\pi \sigma(\rr'')}
  \delta(|\mathbf{r'}-\rr''| - \sigma(\rr'')/2) \frac{\rr'-\rr''}{|\rr'-\rr''|}\\
  \frac{\delta \mathbf{n}_{V1}^A(\rr')}{\delta \sigma(\rr)}
  &= \frac{n(\rr)}{4\pi
    \sigma(\rr)}\delta'(|\rr'-\rr| - \sigma(\rr)/2) \frac{\rr-\rr'}{|\rr-\rr'|}
  +
  \frac{n(\rr)}{2\pi
    \sigma(\rr)^2}\delta(|\rr'-\rr| - \sigma(\rr)/2) \frac{\rr-\rr'}{|\rr-\rr'|}
\end{align}
This completes our exploration of the partial derivatives of weighted
densities of the $A$ variety.

@@

The derivatation of the asymmetrically averaged correlation function
$g_\sigma^S(\rr)$ requires that we express the free energy in terms of
a set of weighted densities in which the radius of the weighting
kernel is determined according to the position at which the weighted
density is being evaluated.  This is less intuitive than the $A$ case,
since \emph{nothing is there}!  However, in practice it corresponds to
the picture shown in Figure~\ref{fig:contact}, which involves
extending the radius for spheres just at the position touching a
given spot, where we are evaluating the correlation function.  We
define the weighted density $n_3^S$ as:
\begin{align}
  n_3^{S}(\rr') &= \int n(\rr'') \Theta(\left|\rr' - \rr''\right| -\sigma(\rr')/2) d\rr''
\end{align}
In the symmetric case, the functional derivative is strikingly simple:
\begin{align}
   \frac{\delta n_3^{S} (\rr')}{\delta \sigma(\rr)} &=
   \delta(\rr'-\rr) \frac 12 \int n (\rr'') \delta(|\rr' - \rr''| - \sigma(\rr')/2)
   \delta(\rr'-\rr) d\rr''
   \\
   &= \frac 12 \delta(\rr'-\rr) n_2^S(\rr')
\end{align}
The functional derivative does \emph{not} eliminate the integral,
which simply changes from one weighted density to another.  The
corresponding term in the free energy functional, once integrated,
takes a simple form:
\begin{align}
  \int \frac{\delta A_{HS}}{\delta n_3^{S}(\rr')}
  \frac{\delta n_3^{S}(\rr')}{\delta \sigma(\rr)} d\rr'
  &= \frac 12 \int \frac{\delta A_{HS}}{\delta n_3^{S}(\rr')} \delta(\rr'-\rr)
  n_2^S(\rr') d\rr' \\
  &= \frac 12 \frac{\delta A_{HS}}{\delta n_3^{S}(\rr)} n_2(\rr)
\end{align}
which means that we can compute $g_\sigma^S(\rr)$ using only
convolutions of the density, in strong contrast with
$g_\sigma^A(\rr)$, which requires convolutions of derivatives of the
free energy density.

>>>>

We now address the symmetrically averaged correlation function, which
is defined in Equation~\ref{eq:gS}.  This corresponds to the
correlation function averaged for spheres \emph{touching at a given
  point}.  In this case, we conceptually would like to evaluate the
pressure felt by the surface of spheres where that surface is located
at point~$\rr$.  We can approximate this value by assuming that this
pressure will be simply related to the free energy density at
point~$\rr$.  Through math similar to the previous derivations, this
leads to the expression
\begin{align}
  %% p_{HS}^S(\mathbf{r}) &= \frac{1}{n_0(\rr) 2\pi \sigma^2} \frac{\partial
  %%   a_{HS}(\rr)}{\partial \sigma} \label{eq:p_{HS}^S}\\
%  \ncontact(\rr) &= \frac{1}{n_0(\rr) k_BT 2\pi \sigma^2} \frac{\delta
%    A_{HS}}{\delta \sigma(\mathbf{r})} \\
  g_\sigma^S(\rr)% &= \frac{\ncontact(\rr)}{n_0(\rr)} \\
  &= \frac{1}{n_0(\rr)^2}\frac{1}{ 2\pi \sigma^2}
  \frac{\partial \Phi(\rr)}{\partial \sigma} \label{eq:g-S}
\end{align}
where $\Phi(\rr) = \Phi_1(\rr) + \Phi_2(\rr) + \Phi_3(\rr)$ is the
dimensionless free energy density.  This requires that we evaluate the
derivatives of the fundamental measures $n_\alpha(\rr)$ with respect
to diameter.  These derivatives, when performed exact, lead to
derivatives of the $\delta$~function, resulting in a functional gives
poor results.  We find that our $g_\sigma^S$ is considerably improved
by using integration by parts to eliminate derivatives of the
$\delta$~function by approximating the density to be smoothly varying,
which leads to derivatives such as:
\begin{align}
  \frac{\partial n_2(\rr)}{\partial \sigma} &\approx \frac{2}{\sigma}n_2(\rr)
\end{align}
We note that the density at a hard surface is \emph{not} smoothly
varying, but that the removal of these terms improves the functional,
while at the same time reducing its computational cost by avoiding the
need to calculate any additional weighted densities or convolutions.

The final formula for the symmetrically averaged correlation function
under these approximations is:
\begin{align}
  g_\sigma^S &= \text{\textcolor{red}{put formula in here, if it is
      short enough}}
\end{align}

\subsection{Gross's asymmetrically averaged correlation functional}\label{sec:gross}
One approximation for the correlation function is that of
Gross\cite{gross2009density}, which is of the asymmetrically averaged
variety ($g_\sigma^A$):
\begin{align}
  g_\sigma^\text{Gross,A}(\rr) &= \frac{1 - \frac{\pi}{12}\sigma^3n_A(\rr)}{\left(1 -
    \frac{\pi}{6}\sigma^3n_A(\rr)\right)^3}
\end{align}
where $n_A$ is the averaged density defined in Equation~\ref{eq:nA}.
This formula is arrived at by using the density averaged over all
spheres that could be touching a sphere at point~$\rr$ in the
Carnahan-Starling equation for the correlation function at contact,
given in Equation~\ref{eq:cs-g}.

@@

%% \section{Method}

%% We implemented the above functionals as part of the Deft software
%% package~\cite{deft}, which uses a regular Cartesian grid with a grid
%% spacing of $0.05\sigma$.  Convolutions are implemented with fast
%% fourier transforms.  For comparison, we performed Monte Carlo
%% simulations of the hard-sphere fluid.  We performed these simulations
%% in a fixed-number ensemble.

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/walls-10}
  \caption{Density and contact densities at a wall with bulk filling
    fraction of 0.1.}
  \label{fig:walls-10}
\end{figure}

@@

We performed a Monte-Carlo simulation of the hard sphere fluid to
measure the contact value of the correlation function for several
simple inhomogeneous configurations.  For each configuration, we
compute the mean density, and the contact values of the correlation
function, averaged as defined in Equations~\ref{eq:gA} and
\ref{eq:gS}.  We compare these with the functionals presented in
sections~\ref{sec:g-A} to~\ref{sec:yuwu}.  We constructed our
functionals using both the original White Bear
functional~\cite{roth2002whitebear} as well as the mark II version of
the White Bear functional~\cite{hansen2006density}, but the results
were essentially indistinguishable on our plots, so we exclusively
show the results due to the original White Bear functional.

We simulate the inhomogeneous hard sphere fluid at four hard-wall
interfaces.  The first and simplest is a simple flat hard wall.  We
then study two convex hard surfaces.  One is an excluded sphere with
diameter $2\sigma$, which corresponds to a ``test particle''
simulation with one of a hard sphere at the origin with diameter
$\sigma$.  The second is an excluded sphere with diameter $6\sigma$,
which demonstrates behavior typical of mildly convex hard surfaces.
Finally, we study a concave surface given by a hard cavity in which
our fluid is free to move up to a diameter of $16\sigma$, which
demonstrates behavior typical of mildly concave surfaces.  In each
case, we performed a low-density (filling fraction 0.1) and high-density
(filling fraction 0.4) simulation.  We performed additional
computations at over a wider range of curvatures and densities, but
chose these as typical examples.

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/inner-4-10}
  \caption{Density and contact densities at a xxx with bulk filling
    fraction of 0.1.}
  \label{fig:inner-4-10}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/inner-12-10}
  \caption{Density and contact densities at a xxx with bulk filling
    fraction of 0.1.}
  \label{fig:inner-12-10}
\end{figure}

@@

We begin by presenting our low-density results, corresponding to a
filling fraction of 0.1, which are shown in
Figures~\ref{fig:walls-10}-\ref{fig:outer-10}.  At this low density,
the contact value of the correlation function in the bulk is only 1.3,
indicating that correlations are indeed small, the fluid should be
relatively easy to model.  Indeed, the contact density at the hard
surface is only around 50\% higher than the bulk, and the FMT
predicted density is close to indistinguishable from the true
density for each of the four configurations.

The $g_\sigma^A$ correlation function in each configuration is very
flat, with only small, smooth changes as the surface is approached.
Our functional $g_\sigma^A$ very closely matches the Monte Carlo
predictions in each case, while that of Gross consistently
underestimates the correlation at the interface by about
\textcolor{red}{few percent}.

The $g_\sigma^S$ correlation function shows considerably more
structure, as well as additional variation due to the curvature of the
hard surface.

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/outer-16-01}
  \caption{Density and contact densities at a xxx with bulk filling
    fraction of 0.1.}
  \label{fig:outer-10}
\end{figure}

@@

We must therefore compute both the $\frac{\delta A_{HS}}{\delta
  n(\mathbf{r}')}$, etc., terms and the $\frac{\delta
  n(\mathbf{r}')}{\delta \sigma(\mathbf{r})}$, etc., terms.  We will
compute the $\frac{\delta A_{HS}}{\delta n(\mathbf{r}')}$, etc., terms
for both the functional derived from the White Bear
functional\cite{roth2002whitebear} and from the White Bear mark
II functional\cite{hansen2006density}.  Following this we will compute
the $\frac{\delta n(\mathbf{r}')}{\delta \sigma(\mathbf{r})}$, etc.,
terms for both the $A$ and $S$ cases.  These latter derivitives will
differ for the $A$ and $S$ cases according to the qualitive concepts
described above.

@@

The derivatation of the asymmetrically averaged correlation function
$g_\sigma^A(\rr)$ requires that we express the free energy in terms of
a set of weighted densities in which the radius of the weighting
kernel is determined according to the position of the sphere in
question---which is to say, the location where the density itself is
evaluated.  This is the most natural way to construct such a
functional, in which the radius of spheres is a function of position.
Thus we define $n_3^A$ as
\begin{align}
  n_3^{A}(\rr') &= \int n(\rr'') \Theta(\sigma(\rr'')/2 -\left|\rr' - \rr''\right|) d\rr''
\end{align}
and we define the other weighted densities similarly.  In order to
compute derivatives of the free energy with respect to $R(\rr)$, we
need to find $g_\sigma^A$.  In the case of $n_3^A$, we need
\begin{align}
  \frac{\delta n_3^{A} (\rr')}{\delta \sigma(\rr)} &=
  \frac 12 \int n (\rr'') \delta(\sigma(\rr'')/2 -\left|\rr' - \rr''\right|) \delta(\rr-\rr'') d\rr'' \\
   &= n (\rr) \delta(\sigma(\rr)/2 -\left|\rr' - \rr\right|)
\end{align}
Similarly, for $n_2^A$, we find
\begin{align}
  n_2^{A}(\rr') &= \int n(\rr'') \delta(\sigma(\rr'')/2 -\left|\rr' - \rr''\right|) d \rr''\\
  \frac{\delta n_2^{A}(\rr')}{\delta \sigma(\rr)} &= \frac 12 n(\rr) \delta'(\sigma(\rr)/2 -\left|\rr' - \rr\right|)\\
\end{align}
This is somewhat more challenging, since it involves an unfamiliar
function, the derivative the Dirac delta function $\delta'$.
%
The derivatives of the remaining scalar densities can be reduced to
sums of the terms above:
\begin{align}
  n_1(\rr') &= \int \mathbf{dr''} \frac{n(\rr'')}{2\pi \sigma(\rr'')}
  \delta(\sigma(\rr'')/2 -\left|\rr' - \rr''\right|) \\
  \frac{\delta n_1(\rr')}{\delta \sigma(\rr)}
  &= \frac{n(\rr)}{4\pi
    \sigma(\rr)}\delta'(\sigma(\rr)/2 -\left|\rr' - \rr\right|)
  -
  \frac{n(\rr)}{2\pi
    \sigma(\rr)^2}\delta(\sigma(\rr)/2 -\left|\rr' - \rr\right|)
\end{align}
and
\begin{align}
  n_0(\rr') &= \int \mathbf{dr''} \frac{n(\rr'')}{\pi \sigma(\rr'')^2}
  \delta(\sigma(\rr'')/2 -\left|\rr' - \rr''\right|) \\
  \frac{\delta n_0(\rr')}{\delta \sigma(\rr)}
  &= \frac{n(\rr)}{2\pi
    \sigma(\rr)^2}\delta'(\sigma(\rr)/2 -\left|\rr' - \rr\right|)
  -
  2 \frac{n(\rr)}{\pi
    \sigma(\rr)^3}\delta(\sigma(\rr)/2 -\left|\rr' - \rr\right|)
\end{align}
The vector-weighted densities have no additional occurrences of $R$,
their functional derivatives come out similarly:
\begin{align}
  \mathbf{n}_{V2}^{A}(\rr') &= \int n(\rr') \delta(\sigma(\rr'')/2 -\left|\rr' - \rr''\right|)
    \frac{\rr'-\rr''}{|\rr'-\rr''|} d \rr''\\
  \frac{\delta \mathbf{n}_{V2}^{A}(\rr')}{\delta \sigma(\rr)} &= -\frac 12 n(\rr) \delta'(\sigma(\rr)/2 -\left|\rr' - \rr\right|)
    \frac{\rr-\rr'}{|\rr-\rr'|}\\
\end{align}
and analogously for $\mathbf{n}_{V1}^A$:
\begin{align}
  \mathbf{n}_{V1}^A(\rr') &= \int d\rr'' \frac{n(\rr'')}{2\pi \sigma(\rr'')}
  \delta(\sigma(\rr'')/2 -\left|\rr' - \rr''\right|) \frac{\rr'-\rr''}{|\rr'-\rr''|}\\
  \frac{\delta \mathbf{n}_{V1}^A(\rr')}{\delta \sigma(\rr)}
  &= -\frac{n(\rr)}{4\pi
    \sigma(\rr)}\delta'(\sigma(\rr)/2 -\left|\rr' - \rr\right|) \frac{\rr-\rr'}{|\rr-\rr'|}
  +
  \frac{n(\rr)}{2\pi
    \sigma(\rr)^2}\delta(\sigma(\rr)/2 -\left|\rr' - \rr\right|) \frac{\rr-\rr'}{|\rr-\rr'|}
\end{align}
This completes our exploration of the partial derivatives of weighted
densities of the $A$ variety.

\end{widetext}
\end{document}

@@

The derivatation of the asymmetrically averaged correlation function
$g_\sigma^S(\rr)$ requires that we express the free energy in terms of
a set of weighted densities in which the radius of the weighting
kernel is determined according to the position at which the weighted
density is being evaluated.  This is less intuitive than the $A$ case,
since \emph{nothing is there}!  However, in practice it corresponds to
the picture shown in Figure~\ref{fig:contact}, which involves
extending the radius for spheres just at the position touching a
given spot, where we are evaluating the correlation function.  We
define the weighted density $n_3^S$ as:
\begin{align}
  n_3^{S}(\rr') &= \int n(\rr'') \Theta(\sigma(\rr'')/2 -\left|\rr' - \rr''\right|) d\rr''
\end{align}
In the symmetric case, the functional derivative is strikingly simple:
\begin{align}
   \frac{\delta n_3^{S} (\rr')}{\delta \sigma(\rr)} &=
   \delta(\rr'-\rr) \frac 12 \int n (\rr'') \delta(\sigma(\rr'')/2 -\left|\rr' - \rr''\right|)
   \delta(\rr'-\rr) d\rr''
   \\
   &= \frac 12 \delta(\rr'-\rr) n_2^S(\rr')
\end{align}
The functional derivative does \emph{not} eliminate the integral,
which simply changes from one weighted density to another.  The
corresponding term in the free energy functional, once integrated,
takes a simple form:
\begin{align}
  \int \frac{\delta A_{HS}}{\delta n_3^{S}(\rr')}
  \frac{\delta n_3^{S}(\rr')}{\delta \sigma(\rr)} d\rr'
  &= \frac 12 \int \frac{\delta A_{HS}}{\delta n_3^{S}(\rr')} \delta(\rr'-\rr)
  n_2^S(\rr') d\rr' \\
  &= \frac 12 \frac{\delta A_{HS}}{\delta n_3^{S}(\rr)} n_2(\rr)
\end{align}
which means that we can compute $g_\sigma^S(\rr)$ using only
convolutions of the density, in strong contrast with
$g_\sigma^A(\rr)$, which requires convolutions of derivatives of the
free energy density.

The functional derivative of $n_2^S(\rr')$ is quite similar:
\begin{align}
  n_2^{S}(\rr') &= \int n(\rr'') \delta(\sigma(\rr'')/2 -\left|\rr' - \rr''\right|)d\mathbf
  r''\\
  \frac{\delta n_2^{S}(\rr')}{\delta \sigma(\rr)} &= \delta(\rr-\rr') \frac 12 \int n(\rr'')
  \delta'(\sigma(\rr'')/2 -\left|\rr' - \rr''\right|) d\rr''
  \\
  &= \delta(\rr-\rr') \frac12 n_{2'}(\rr')
\end{align}
where we recognize that the convolution---while not one of the standard
FMT weighted density---is a simple convolution of the density, which
we define in Equation~\ref{n2p} as $n_{2'}$.  We can
go on to compute the derivatives of the remaining weighted densities:
\begin{align}
  n_1^{S}(\rr') &= \int \frac{n(\rr'')}{2\pi \sigma(\rr')} \delta(\sigma(\rr'')/2 -\left|\rr' - \rr''\right|)d\mathbf r''\\
    \frac{\delta n_1^{S}(\rr')}{\delta \sigma(\rr)} &=
    \delta(\rr-\rr') \left( \int \frac{n(\rr'')}{4\pi \sigma(\rr)}
    \delta'(\sigma(\rr'')/2 -\left|\rr' - \rr''\right|) d\rr''
    - \frac 12 n_0^s(\rr') \right)
    \\
    &= \delta(\rr-\rr')\left(
    \frac{n_{2'}(\rr')}{4\pi\sigma} - \frac12n_0(\rr')
    \right)
\end{align}
\begin{align}
  n_0^{S}(\rr') &= \int \frac{n(\rr'')}{\pi \sigma(\rr')^2} \delta(\sigma(\rr'')/2 -\left|\rr' - \rr''\right|)d\mathbf r''\\
    \frac{\delta n_0^{S}(\rr')}{\delta \sigma(\rr)}
    &= \delta(\rr-\rr') \left( \frac 12 \int \frac{n(\rr'')}{\pi \sigma(\rr)^2}
    \delta'(\sigma(\rr'')/2 -\left|\rr' - \rr''\right|) d\rr''
    -
    \frac{2}{\sigma} n_0^S(\rr') \right)
    \\
    &= \delta(\rr-\rr')\left(
    \frac{n_{2'}(\rr')}{2\pi\sigma^2}
    - \frac{2}{\sigma} n_0(\rr')
    \right)
\end{align}
\begin{align}
  \mathbf{n}_{V2}^{S}(\rr') &= \int n(\rr'') \delta(\sigma(\rr'')/2 -\left|\rr' - \rr''\right|)
  \frac{(\rr' - \rr'')}{|\rr' - \rr''|} d \rr''\\
  \frac{\delta \mathbf{n}_{V2}^{S}(\rr')}{\delta \sigma(\rr)} &= \delta(\rr - \rr')
  \frac 12 \int n(\rr'') \delta'(\sigma(\rr'')/2 -\left|\rr' - \rr''\right|)
  \frac{(\rr - \rr'')}{|\rr - \rr''|} d\rr''
  \\
  &= \delta(\rr-\rr')
   \frac12\mathbf{n}_{V2'}(\rr')
\end{align}
\begin{align}
 \mathbf{n}_{V1}^{S}(\rr') &= \int \frac{n(\rr'')}{2\pi \sigma(\rr')} \delta(\sigma(\rr'')/2 -\left|\rr' - \rr''\right|)d\mathbf r''
   \frac{(\rr' - \rr'')}{|\rr' - \rr''|}\\
 \frac{\delta \mathbf{n}_{V1}^{S}(\rr')}{\delta \sigma(\rr)}
 &= \delta(\rr-\rr')\left(\int \frac{n(\rr'')}{4\pi \sigma(\rr)}
   \delta'(\sigma(\rr'')/2 -\left|\rr' - \rr''\right|) \frac{(\rr - \rr'')}{|\rr - \rr''|}  d\rr''
   - \frac{\mathbf{n}_{V2}^S(\rr')}{2 \pi \sigma(\rr)^2} \right)
   \\
   &= \delta(\rr-\rr')\left(
   \frac{\mathbf{n}_{V2'}(\rr')}{4\pi \sigma}
   - \frac{\mathbf{n}_{V2}(\rr')}{2 \pi \sigma^2}
   \right)
\end{align}
